Explainable AI with Probabilistic Graphical Models

### 2019-08-28 mathematical-background/graph-theory.tex
Capitoli completi

\section{Graph Theory} \label{sec:graph-theory}
Many problems in Machine Learning (ML) don't involve classification or prediction of single data points in isolation, but of set of entities that may present a more, or less, complex relation with each other. 
Most real-world phenomena fit into the latter framework.
Graphs are one of the most powerful tools for the modelling of this class of problems, as their structure naturally captures the wide variety of relations that may exist between entities.
These range from the atomical structure of a molecule to a social network of friends.  
In both these examples graphs help in reasoning, visualising and making inferences and predictions.

\subsection{Graphs} \label{subsec:graphs}
\begin{definition}
	A graph is a tuple 
	\begin{equation}
	\mathcal{G} = (\mathcal{V}, \mathcal{E})
\end{equation}
with $\mathcal{V} = \{ v_1 \ldots v_n \}$ the set of \textit{vertices} and $\mathcal{E} = \mathcal{V} \times \mathcal{V}$ the set of \textit{edges}.
\end{definition}

For our scopes, we will only be considering the case where every element in $\mathcal{E}$ is a pair either of the form $(v_i, v_j)$ or $(v_j, v_i)$ with $i \neq j$.  
That is to say that the class of graphs presently of interest for us are those where there can be at most a single directed edge between any node in $\mathcal{V}$ and no self-loops.
We are also interested in enforcing that there be no \textit{cycles} in the graph, i.e. sequences of nodes of the form $v_i \rightarrow v_j \rightarrow \cdots \rightarrow v_i$.
The resulting graph possessing only directed edges and no cycles is commonly called a \textit{directed acyclic graph}, or DAG for short.  
This data structure is of paramount importance as it's the fundamental graphical representation used for Bayesian Networks.

An example of a DAG, containing five nodes, is shown in Fig. \ref{fig:bn-example-dag}.

\begin{figure}[htbp]
\centerline{\includegraphics[width=\columnwidth]{mathematical-background/images/bn-example-structure}}
\caption{Example DAG representing a subset of the data set used in this thesis}
\label{fig:bn-example-dag}
\end{figure}

\subsection{Polytrees} \label{subsec:polytrees}
We now have all elements to be able to formally define a Bayesian Network.
I will also define polytrees and trees because these are a fundamental concept for the work carried out in this thesis.
\begin{definition}
	A \textit{loop} is a trace $v_i, v_j \ldots v_i$ of nodes obtained by following edges regardless of their direction
\end{definition}
\begin{definition}
	A directed graph containing no such loops is called a \textit{polytree}. 
\end{definition}
\begin{definition}
	A \textit{tree} is a particular case of polytree where each node has at most one parent.	
\end{definition}

\subsection{D-separation} \label{subsec:d-separation}
\textit{Dependence-separation} or \textit{d-separation}, as the name entails, is a concept relating to the conditional dependence between variables.
It was first presented by \cite{Pearl1988}
To define it, we first have to clarify when two sets of nodes $X$ and $Y$ are causally connected.
This is so if $Z = \emptyset$ and they are part of one of the following three structures, called \textit{v-structures} in this context:
\begin{itemize}
  \item $X \rightarrow Z \rightarrow Y$
  \item $X \leftarrow Z \leftarrow Y$
  \item $X \leftarrow Z \rightarrow Y$
\end{itemize}
This means that knowing something about $X$ also tells us something new about $Y$.
$X$ and $Y$ are causally independent if they appear in the following v-structure:
\begin{itemize}
  \item $X \rightarrow Z \leftarrow  Y$
\end{itemize}
Such a configuration is called a \textit{collider} and it blocks the flow of information from $X$ to $Y$.
If $Z \neq \emptyset$ the cases are reversed so colliders are open and the other three structures are blocked.
\begin{definition}
	Given disjoint subsets $X, Y, Z \subset \mathcal{X}$, $X$ and $Y$ are d-separated if:
	\begin{itemize}
		\item $Z \neq \emptyset$: no path between $X$ and $Y$ presents a collider
		\item $Z = \emptyset$: there is a collider on every path between $X$ and $Y$
	\end{itemize}
\end{definition}

The independencies between variables are encoded in the structure of the DAG so every distribution whose BN has the same connections between nodes also has the same independencies, regardless of the values of the variables.

A series of examples using the DAG presented in Fig. \ref{fig:bn-example-dag} are shown in Fig. \ref{fig:bn-separations-example-1}, \ref{fig:bn-separations-example-2}, \ref{fig:bn-separations-example-3}.
We can see how the network's topology and the nodes chosen to be in the observed set $Z$, define the resulting separations.
In all cases $X= \{ \text{eta arrotondata} \}$ and $Y=V \smallsetminus X \smallsetminus Z$; we are asking for the set of all nodes in the DAG that are d-separated from $X$, given evidence $Z$.
The reason for this can easily be done by enumerating all paths through the v-structures in the network and applying the definitions for causal connections given above.
In the case shown in Fig. \ref{fig:bn-separations-example-1} we see that the node \textbf{eta arrotondata} is separated from nodes \textbf{recettori estrogeni}, \textbf{differenziazione} and \textbf{pN} given the observed evidence \textbf{mut17q21}.
The reason for this is because \textbf{$\text{eta arrotondata} \leftarrow \text{mut17q21} \rightarrow \text{recettori estrogeni}$} is a \textit{fork} and thus the flow of information from the rest of the network is blocked.
The way in which changing the conditioning set $Z$ also changes the independencies, can clearly be seen in Fig. \ref{fig:bn-separations-example-2} and \ref{fig:bn-separations-example-3}.

\begin{figure}[htbp]
\centerline{\includegraphics[width=\columnwidth]{mathematical-background/images/bn-example-separations-1}}
\caption{D-Separations in a subset of the provided data set (see Sec. \ref{sec:data-set})}
\label{fig:bn-separations-example-1}
\end{figure}

\begin{figure}[htbp]
\centerline{\includegraphics[width=\columnwidth]{mathematical-background/images/bn-example-separations-2}}
\caption{D-Separations in a subset of the provided data set (see Sec. \ref{sec:data-set})}
\label{fig:bn-separations-example-2}
\end{figure}

\begin{figure}[htbp]
\centerline{\includegraphics[width=\columnwidth]{mathematical-background/images/bn-example-separations-3}}
\caption{D-Separations in a subset of the provided data set (see Sec. \ref{sec:data-set})}
\label{fig:bn-separations-example-3}
\end{figure}

>>>>

\section{Graph Theory} \label{sec:graph-theory}
Many problems in Machine Learning (ML) do not involve classification or prediction of single data points in isolation, but of sets of entities that may present a more, or less, complex relation with each other. 
Most real-world phenomena fit into the latter framework.
Graphs are one of the most powerful tools for the modelling of this class of problems, as their structure naturally captures the wide variety of relations that may exist between entities.
These range from the atomic structure of a molecule to a social network of friends.  
In both these examples graphs help in reasoning, visualising and making inferences and predictions.

\subsection{Graphs} \label{subsec:graphs}
\begin{definition}[Directed Graph]
	A directed graph is a tuple 
	\begin{equation*}
		\mathcal{G} = (\mathcal{V}, \mathcal{E}) \,,
	\end{equation*}
with $\mathcal{V} = \{ v_1 \ldots v_n \}$ the set of vertices/nodes and $\mathcal{E}\subseteq \mathcal{V} \times \mathcal{V}$ the set of edges.
\end{definition}
We will not be needing the subclass known as \textit{undirected graphs} that are characterised by $\mathcal{E}$ being a \textit{set} of unordered pairs, that is of sets of the form $\{x,y\}$, with $x,y \in \mathcal{V}$.

The class of graphs presently of interest are those where there can be at most a single directed edge between any pair of nodes in $\mathcal{V}$; that is, we are not considering \textit{multigraphs}.
We are also interested in enforcing that there be no \textit{cycles} in the graph so there can be no subset of edges in $\mathcal{E}$ that when followed from vertex $v_i$ eventually ends up in $v_i$ again.
A cycle is a \textit{walk} - a sequence of edges which joins a sequence of vertices - of nodes of the form $v_i, v_j, \cdots, v_i$ i.e., a walk where only the first and last vertex are repeated.
Thus we have also automatically excluded the special case of cycle called \textit{self-loop}: an edge from a node to itself. 
The resulting graph possessing only directed edges and no cycles is commonly called a \textit{directed acyclic graph}, or DAG for short.  
\begin{definition}[Directed Acyclic Graph] \label{def:dag}
	A directed acyclic graph is a graph where every edge is directed and there are no cycles.
\end{definition}

In a DAG we may qualify nodes based on their \enquote{relationship status}:
\begin{description}
	\item[children] The children of node $u$ are all nodes $k$ for which there is a \textit{directed edge} from $u$ to $k$
	\item[parents] The parents of node $u$ are all nodes $k$ for which there is a \textit{directed edge} from $k$ to $u$
	\item[descendants] The descendants of node $u$ are all nodes $k$ for which there is a \textit{directed path} i.e., a walk where all vertices are distinct, from $u$ to $k$
	\item[ancestors] The ancestors of $u$ are all nodes for which there is a directed path from $k$ to $u$
\end{description}

An example of a DAG, containing five nodes, is shown in Figure \ref{fig:bn-example-dag}.

\begin{figure}[htbp]
\centerline{\includegraphics[width=0.5\textwidth]{mathematical-background/images/bn-example-structure}}
\caption{Example DAG representing a subset of the data set used in this thesis}
\label{fig:bn-example-dag}
\end{figure}

Polytrees and trees will also be defined because these are a fundamental concept for the work carried out in this thesis.
\begin{definition}[Tree] \label{def:tree}
	A tree is an undirected graph where there is one and only one walk between every node.	
\end{definition}
\begin{definition}[Polytree] \label{def:polytree}
	A polytree is a DAG whose underlying undirected graph is a tree.
	That is, if the directionality of edges is removed from the DAG, the resulting object is a tree. 
\end{definition}

\subsection{D-separation} \label{subsec:d-separation}
\textit{Dependence-separation} or \textit{d-separation}, as the name entails, is a concept relating to the conditional dependence between variables and was first presented by \citet{Pearl1988}.
We define the notation $u \rightarrow v$ to signify that there is a \textit{trail} between $u$ and $v$ in the graph, with a trail $(u, \ldots ,v)$ being a walk where all edges are distinct.

$u$ and $v$ and a node $z \in Z$ may be arranged in the graph in one of the following four configurations, called \textit{v-structures} in this context:
\begin{itemize}
  \item \textit{chain}: $u \rightarrow z \rightarrow v$
  \item \textit{chain}: $u \leftarrow z \leftarrow v$
  \item \textit{fork}: $u \leftarrow z \rightarrow v$
  \item \textit{collider}: $u \rightarrow z \leftarrow v$
\end{itemize}

These v-structures may be \textit{closed} by the set $Z$, this happens when:
\begin{itemize}
  \item \textit{chain}: $u \rightarrow z \rightarrow v$ and $z \in Z$
  \item \textit{chain}: $u \leftarrow z \leftarrow v$ and $z \in Z$
  \item \textit{fork}: $u \leftarrow z \rightarrow v$ and $z \in Z$
  \item \textit{collider}: $u \rightarrow z \leftarrow v$ and $z \notin Z$ and no descendant $z'$ of $z$ i.e., $z'$ such that $z \rightarrow z'$ exists, is also in the set $Z$
\end{itemize}

We say that $u$ and $v$ are \textit{d-separated} by $Z$ if every v-structure they appear in, is closed by $Z$.
Conversely, if there is at least one \textit{open} v-structure $u$ and $v$ are \textit{d-connected}.

If $u$ and $v$ are d-connected, then knowing something about $u$ also tells us something new about $v$, and viceversa.
An intuition for this can be given by interpreting the paths \textit{causally}.
In the case of a \textit{chain}, $z$ is the cause of $v$ so knowing $z$ tells us everything we need to know about the value of $v$ (or of $u$, if the chain is reversed).
In a \textit{fork}, conditioning on the \textit{common cause} $z$ has the same effect: $z$ is sufficient to know $u$ and $v$.
This is also called the \enquote{Common Cause Principle} \citep{sober1988principle}.

A good intuition for when $u$ and $v$ are \textit{d-separated} was given by \citet{Pearl1988}: imagine that there are two independent causes for a car refusing to start ($z$): having no gas ($u$) and having a dead battery ($v$): $u \rightarrow z \leftarrow  v$.
Only knowing that the battery is charged gives no information about the car having fuel or not.
But if we now know that the battery is charged after observing that the car won't start, we know for sure that it must be out of fuel.
So knowing something about $u$ is informative about $v$, after conditioning on $z$.

\begin{definition}[D-Separation] \label{def:d-separation}
	Given vertices $u$ and $v$ and a set of vertices $Z$, then $u$ and $v$ are d-separated by $Z$ if:
	\begin{itemize}
		\item $Z \neq \emptyset$ and $u$ and $v$ are never part of a collider;
		\item $Z = \emptyset$ and $u$ and $v$ are always part of a collider.
	\end{itemize}
\end{definition}

The independencies between variables are encoded in the structure of the DAG so every probability distribution modelled by a BN that has the same connections between nodes, also necessarily has the same independencies regardless of the values of the variables.

A series of examples using the DAG presented in Figure \ref{fig:bn-example-dag} are shown in Figures \ref{fig:bn-separations-example-1}, \ref{fig:bn-separations-example-2}, \ref{fig:bn-separations-example-3}.
We can see how the network's topology and the nodes chosen to be in the observed set $Z$, define the resulting separations.
In all cases $u=  \text{eta arrotondata} $ and $Y=V \mysetminus u \mysetminus Z$; we are asking for the set of all nodes in the DAG that are d-separated from $u$, given evidence $Z$.
This can easily be answered by enumerating all the v-structures in the network and applying Definition \ref{def:d-separation}.
In the case shown in Figure \ref{fig:bn-separations-example-1} we see that the node \textbf{eta arrotondata} is separated from nodes \textbf{recettori estrogeni}, \textbf{differenziazione} and \textbf{pN} given the observed evidence \textbf{mut17q21}.
The reason for this is because \textbf{$\text{eta arrotondata} \leftarrow \text{mut17q21} \rightarrow \text{recettori estrogeni}$} is a \textit{fork} and thus the flow of information from the rest of the network is blocked.
The way in which changing the conditioning set $Z$ also changes the independencies, can clearly be seen by comparing Figures \ref{fig:bn-separations-example-2} and \ref{fig:bn-separations-example-3}.

\begin{figure}[htbp]
\centerline{\includegraphics[width=0.5\textwidth]{mathematical-background/images/bn-example-separations-1}}
\caption{D-Separations in a subset of the provided data set (see Section \ref{sec:data-set})}
\label{fig:bn-separations-example-1}
\end{figure}

\begin{figure}[htbp]
\centerline{\includegraphics[width=0.5\textwidth]{mathematical-background/images/bn-example-separations-2}}
\caption{D-Separations in a subset of the provided data set (see Section \ref{sec:data-set})}
\label{fig:bn-separations-example-2}
\end{figure}

\begin{figure}[htbp]
\centerline{\includegraphics[width=0.5\textwidth]{mathematical-background/images/bn-example-separations-3}}
\caption{D-Separations in a subset of the provided data set (see Section \ref{sec:data-set})}
\label{fig:bn-separations-example-3}
\end{figure}

### 2019-08-28 mathematical-background/information-theory.tex
Literature Review
Capitoli completi

\section{Information Theory} \label{sec:information-theory}
The birth of the field of \textit{information theory} is usually traced back to the seminal paper \enquote{A Mathematical Theory of Communication} (\cite{Shannon1949}) where Claude Shannon set the mathematical basis for the quantification of the amount of \textit{information} transmissible over a noisy channel. 
In his words \enquote{The fundamental problem of communication is that of reproducing at one point, either exactly or approximately, a message selected at another point.}
The concepts of field are broad enough to have influenced practically every other scientific discipline and deep enough to have enabled the \enquote{digital age}, for example by enabling the creation of ever more complicated coding schemes for the compression, reconstruction and obfuscation of digital data.

\subsection{Entropy} \label{subsec:entropy}
In classical mechanical statistics, entropy can be seen as a measure of the uncertainty, or randomness, of a physical system.  
This concept was reapplied by Shannon to measure the amount of randomness in a random variable.
\begin{definition}
	Given a random variable $X$ with probability distribution $\mathbb{P}(X)$, its entropy $\Eta(X)$ is defined as the expected amount of information content carried by $X$ (\cite{Schneider2005}):
\begin{equation} \label{eq:entropy}
	\Eta(X) = \mathbb{E}(I(X)) = \mathbb{E}(-\log (\mathbb{P}(X)) = -\sum_{i=1}^{n} \mathrm{P}\left(x_{i}\right) \log _{b} \mathrm{P}\left(x_{i}\right)
\end{equation}
\end{definition}
The base $b$ of the logarithm defines the unit of measure.  Shannon used $b=2$ as he was dealing with the transmission of digital, binary-coded data; in this case the unit of measure are $bits$.

The simplest example of how information entropy characterises a random variable $X$, is in imagining $X$ to model a coin and the task being to predict the probability of the outcome of a throw being heads.
If the coin is fair, we will not be any more surprised to see the outcome being heads than tails; the entropy is maximum as there is maximum uncertainty regarding the outcome.
However, if the coin is not fair and tails is more probable the we will be more surprised than not to see the outcome being heads.  
The entropy is sub-maximal because there is less uncertainty regarding the outcome: tails is more probable than heads.
If one of the outcomes is impossible, for example if the coin has two heads, then the entropy of the coin is $0$ as there is no uncertainty regarding the result of a toss.

\subsection{Normalised Entropy} \label{subsec:normalised-entropy}
Plain entropy is not a good choice when trying to characterise random variables with different cardinalities of their sample space.
Let us suppose that the objective is to find the variable with the least \enquote{entropic} distribution and we suppose that their values have all been generated by the same process, say Gaussian.
Simply calculating their entropies and ordering them according to this criterion will bias the selection process towards the variables with smallest cardinality.
This is because we supposed them to be distributed in the same way so there will naturally be less uncertainty when there are fewer possible outcomes.
This can easily be understood by imagining the distributions to all be random uniform.

To obviate to this problem we need to \textit{normalise} the entropy so that different-sized variables can be directly compared to each other.
To achieve this, we can look at a measure of \textit{normalised entropy} or \textit{efficiency}:
\begin{equation} \label{eq:normalisedentropy}
 	\eta(X)=-\sum_{i=1}^{n} \frac{p\left(x_{i}\right) \log _{b}\left(p\left(x_{i}\right)\right)}{\log _{b}(n)}
\end{equation}
From Eq. \ref{eq:normalisedentropy} it can be seen that $\eta(X) \in [0,1]$; it is thus normalised and comparable among distributions.
This ratio expresses the amount of entropy found in the distribution compared to the maximum possible entropy when using $n$ symbols, corresponding to the uniform distribution:
\begin{equation}
\mathrm{H}\left(\underbrace{\frac{1}{n}, \ldots, \frac{1}{n}}_{n}\right) = - \sum_{i=1}^n \frac{1}{n} \log _{b} \left( \frac{1}{n} \right) = -n \cdot \frac{1}{n} \log _{b} \left( \frac{1}{n} \right) = - \log _{b} \left( \frac{1}{n} \right) = \log _{b}(n) 
\end{equation}

>>>>

\section{Information Theory} \label{sec:information-theory}
The birth of the field of \textit{information theory} is usually traced back to the seminal paper \enquote{A Mathematical Theory of Communication} \citep{Shannon1949} where the mathematical basis for the quantification of the amount of \textit{information} transmissible over a noisy channel was set. 
In his words, \enquote{the fundamental problem of communication is that of reproducing at one point, either exactly or approximately, a message selected at another point}.
The concepts of field are broad enough to have influenced practically every other scientific discipline and deep enough to have made the \enquote{digital age} possible, for example by enabling the creation of ever more complicated coding schemes for the compression, reconstruction and obfuscation of digital data.

\subsection{Entropy} \label{subsec:entropy}
In classical mechanical statistics, \textit{entropy} can be seen as a measure of the uncertainty, or randomness, of a physical system.  
This concept was reapplied by Shannon to measure the amount of randomness, the converse of information, in a random variable.
\begin{definition}[Entropy] \label{def:entropy}
	The entropy $\Eta(X)$ is defined as the expected amount of information content carried by random variable $X$ \citep{Schneider2005}:
\begin{equation*}
	\Eta(X) = -\sum\limits_{x \in \mathcal{X}} \mathbb{P}(X=x) \log_{b} \mathbb{P}(X=x) \,.
\end{equation*}
\end{definition}
The base $b$ of the logarithm defines the unit of measure.  
Shannon used $b=2$ as he was dealing with the transmission of digital, binary-coded data; in this case the unit of measure would be \textit{bits}.

The entropy of a generic random variable's probability mass function is a \textit{unimodal functional} (a function whose codomain is in the space of functions is called a functional) whose domain is the subset $[0,1]$.
Its maximum $1$ is reached when applied to a uniform probability mass function while its minima appear in the presence of \textit{degenerate probability mass distributions} i.e., those that are localised at a single value (all values have probability zero, except one that is certain).

As always, we can define a \textit{conditional version} of the notion:
\begin{definition}[Conditional Entropy] \label{def:conditional-entropy}
	The conditional entropy $\Eta(Y \mid X)$ is defined as:
\begin{equation*}
	\Eta(Y | X)=-\sum_{x \in \mathcal{X}, y \in \mathcal{Y}} \mathbb{P}(X=x,Y=y) \log_{b} \frac{\mathbb{P}(X=x,Y=y)}{\mathbb{P}(X=x)} \,.
\end{equation*}
\end{definition}

The simplest example of how information entropy characterises a random variable $X$, is in imagining $X$ as modelling a coin and being tasked with predicting the probability of the outcome of a throw being \textit{heads}.
If the coin is fair, we will not be any more \enquote{surprised} to see that the outcome is heads instead of tails; the entropy is maximum, as there is maximum uncertainty regarding the outcome.
However, if the coin is not fair and tails is more probable than heads, then we will be more surprised if we see that the outcome is heads.  
The entropy is sub-maximal because there is less uncertainty regarding the outcome due to the fact that tails is more probable than heads.
If one of the outcomes is impossible, for example if the coin has two heads, then the entropy of the coin is $0$, as there is no uncertainty regarding the result of a toss.
All these cases can easily be understood by looking at Figure \ref{fig:entropy-plot}.

%\begin{tikzpicture}[declare function = {b(\k, \n, \p) = \n!/(\k!*(\n-\k)!)*\p**\k*(1-\p)**(\n-\k);}]
%\begin{axis}[domain=0:10,
%  samples=100,
%  enlarge x limits=false,
%  grid=both,
%  no markers,
%  axis equal
%  ]
%\addplot +[thick] {sum [k=0:\N] -b(k,\N,x)*log(b(k,\N,x))/log(2)};
%\end{axis}
%\end{tikzpicture}

\todo[inline]{generare io figura}

\begin{figure}[htbp]
\centerline{\includegraphics[width=0.5\textwidth]{mathematical-background/images/entropy-plot}}
\caption{Entropy of a fair coin with the outcome $X=1$ of the random variable representing heads [original work by Brona, published on Commons at Image:Binary entropy plot.png. Converted to SVG by Alessio Damato, CC BY-SA 3.0, https://commons.wikimedia.org/w/index.php?curid=1984868]}
\label{fig:entropy-plot}
\end{figure}

Plain entropy might be not the appropriate tool when trying to characterise random variables with codomains of different cardinality.
Let us suppose that the objective is to find the variable with the least \enquote{entropic} distribution and imagine that their values have all been generated by the same process, say Gaussian.
Simply calculating their entropies and ordering them according to this criterion will bias the selection process towards the variables with smallest cardinality.
This is because we supposed them to be homoscedastic so there will naturally be less uncertainty when there are fewer possible outcomes.
This can easily be understood by imagining the distributions to all be random uniform
(this topic is further analysed in Subsection \ref{subsec:entropy-based-selection}).

To obviate to this problem we need to \textit{normalise} the entropy so that different-sized variables can be directly compared to each other.
To achieve this, we can look at a measure of \textit{normalised entropy}.
\begin{definition}[Normalised Entropy] \label{def:normalisedentropy}
The normalised entropy - or efficiency - or random variable $X$ with domain $\mathcal{X}$ is given by:
	\begin{equation*} 
 	\Eta(X)=-\sum\limits_{x \in \mathcal{X}} \frac{\mathbb{P}(X=x) \log _{b}\mathbb{P}(X=x)}{\log_{b}(n)} \,.
\end{equation*}
\end{definition}
From Definition \ref{def:normalisedentropy} it can be seen that $\Eta(X) \in [0,1]$; it is thus a normalised measure that is comparable across different-sized variables.
This ratio expresses the amount of entropy found in the distribution compared to the maximum possible entropy when using $n$ symbols, the case corresponding to the uniform distribution:
\begin{equation}
\mathrm{H}\left(\underbrace{\frac{1}{n}, \ldots, \frac{1}{n}}_{n}\right) = - \sum\limits_{i=1}^n \frac{1}{n} \log _{b} \left( \frac{1}{n} \right) = -n \cdot \frac{1}{n} \log _{b} \left( \frac{1}{n} \right) = - \log _{b} \left( \frac{1}{n} \right) = \log _{b}(n) 
\end{equation}

\subsection{Mutual Information} \label{subsec:mutual-information}
Another way of characterising the interrelatedness of two variables is through the concept of \textit{mutual information}, as defined in \citet{Cover2006}, that is closely linked to entropy (Definition \ref{def:entropy}).
\begin{definition}[Mutual Information]
	The mutual information of two random variables $X$ and $Y$ with domains $\mathcal{X}$ and $\mathcal{Y}$ is given by:
	\begin{equation*} \label{def:mutual-information}
		I(X,Y) = \sum\limits_{x \in \mathcal{X}} \sum\limits_{y \in \mathcal{Y}} \mathbb{P}(X=x,Y=y) \log_{b} \left( \frac{\mathbb{P}(X=x,Y=y)}{\mathbb{P}(X=x) \mathbb{P}(Y=y)} \right) \,.
	\end{equation*}
\end{definition}
NB: In Information Theory, the convention is that $0 \log(0) = 0$.

$I(X,Y)$, intuitively, measures the amount of information that $X$ and $Y$ share and can also be seen as the degree to which one variable is informative of the other.
If $X$ and $Y$ are independent then they share no mutual information and something about one of the two gives no information about the other.
This can be immediately understood by rewriting Definition \ref{def:mutual-information} as:
\begin{equation}
	I(X,Y) = H(X) - H(X|Y) = H(Y) - H(Y|X)
\end{equation}
The mutual information I(X,Y) is the reduction in uncertainty of one of the variables given the knowledge of the other.
If $X$ and $Y$ are perfectly correlated (correlation $\rho(X,Y)$, as defined by \citet{Stolp2006}, is a measure of the degree to which two random variables are linearly dependent, perfect correlation means $\rho(X,Y)= \pm 1$) then they both convey the same amount of information and $I(X,Y)$ is equal to the entropy i.e., $\Eta(X) = \Eta(Y)$.

\subsection{Kullback-Leibler Divergence} \label{subsec:kl-divergence}
The \textit{Kullback-Leibler divergence} was first defined by \citet{Kingman2007} and is another measure for the difference between two random variables, that is also closely related to the concepts of entropy (Subsection \ref{subsec:entropy}) and of mutual information (Subsection \ref{subsec:mutual-information}).
\begin{definition}[Kullback-Leibler Divergence] \label{def:kl-divergence}
	The Kullback-Leibler divergence $D_{KL}(Y \parallel X)$ - also known as information gain - between two random variables $Y$ and $X$, whose distribution is defined on the same probability space, is given by:
	\begin{equation*} 
		D_{KL}(Y \parallel X)=-\sum\limits_{x \in \mathcal{X}} \mathbb{P}(Y=y) \log_{b} \left(\frac{\mathbb{P}(X=x)}{\mathbb{P}(Y=y)}\right) = \sum\limits_{x \in \mathcal{X}} \mathbb{P}(Y=y) \log_{b} \left(\frac{\mathbb{P}(Y=y)}{\mathbb{P}(X=x)}\right) \,.
	\end{equation*}
\end{definition}
As always, the base $b$ of the logarithm defines the unit of measure; KL divergence is defined for any $b$.

Intuitively, it can be seen as measuring the amount of information gained when revising one's beliefs from the distribution of $X$ to the distribution of $Y$.
Unlike information gain, it is not a \textit{distance measure} as it is evidently asymmetric i.e., $D_{\mathrm{KL}}(Y \parallel X) \neq D_{\mathrm{KL}}(X \parallel Y)$; if there is an \textit{information gain} when moving from $X$ to $Y$, there is an \textit{information loss} when revising one's beliefs from $Y$ to $X$.

The \textit{mutual information} $I(X,Y)$ is related to the \textit{information gain} by:
\begin{equation} \label{def:mutual-information-kl-divergence-equivalence}
	I(X,Y) = D_{\mathrm{KL}}(\mathbb{P}(X,Y) \| \mathbb{P}(X=x) \mathbb{P}(Y) )
\end{equation}
That is, the mutual information of two random variables is equal to the difference in information content between the product of the marginals and the joint distribution.
This can better be understood through the terms of information theory: the mutual information of $X$ and $Y$ is the extra number of symbols needed to discriminate between the distributions of $X$ and $Y$, when they are coded using their marginal distributions and not their joint.
This is to be expected because the definition of mutual information is exactly the information shared by $X$ and $Y$ and all that is needed to discriminate between the marginals is this particular \enquote{set} of information.
The equivalence shown in Equation \ref{def:mutual-information-kl-divergence-equivalence} can be proved by simply substituting $X=\mathbb{P}(X=x) \mathbb{P}(Y)$ and $Y=\mathbb{P}(X,Y)$ into Definition \ref{def:kl-divergence}; the result will be Definition \ref{def:mutual-information}.

\subsection{Hamming Distance} \label{subsec:hamming-distance}
The \textit{Hamming Distance} is a widely-used distance measure that quantifies the similarity of strings.
\begin{definition}[Hamming Distance] \label{def:hamming}
	The Hamming Distance $D_H(x,y)$ between two vectors $x$ and $y$ is given by:
	\begin{equation}
		d_H(x,y) = \sum\limits_i \Gamma(x_i, y_i) \,,
	\end{equation}
	with $\Gamma(i,j)$ defined as:
	\begin{equation}
		\Gamma(i,j) = 
		\begin{cases}
			0 \quad i \neq j \\
			1 \quad i = j	
		\end{cases} \,.
	\end{equation}
\end{definition}
Given strings of characters, or more in general vectors over some field, of equal length, their Hamming Distance is the number of positions where they differ.
It can be seen as the number of substitutions needed to transform one into the other.

This is a valid distance measure because:
\begin{itemize}
  \item it is non-negative: $D_H(x,y) \geq 0$
  \item it fulfils the identity of indiscernibles: if $x = y$ then $D_H(x,y)=0$
  \item it respects the \textit{Triangle Inequality}: $D_H(x,y) \leq D_h(x,z) + D_H(z,y) $
\end{itemize}

As an example, strings $x=01234$ and $y=15244$ have $D_H(x,y)=2$, as they differ in two positions.

\subsection{Jaccard Distance} \label{subsec:jaccard-distance}
The \textit{Jaccard Distance} is a popular metric to measure the similarity of sets.
\begin{definition}[Jaccard Similarity Coefficient] \label{def:jaccard}
	The Jaccard Similarity Coefficient $J(A,B)$ of two sets $A$ and $B$ is given by:
	\begin{equation*}
		J(A,B) = \frac{|A \cap B |}{| A \cup B |} \,.
	\end{equation*}
\end{definition}
\begin{definition}[Jaccard Distance]
	The Jaccard Distance $D_J(A,B)$ between two sets $A$ and $B$ is given by:
	\begin{equation*}
		D_J(A,B) = 1 - J(A,B) \,.
	\end{equation*}
\end{definition}

This is a valid distance measure because:
\begin{itemize}
  \item it is non-negative: $D_J(A,B) \geq 0 \quad \forall A, B$
  \item it fulfils the identity of indiscernibles: if $A = B$ then $D_J(A,B)=0$
  \item it respects the \textit{Triangle Inequality}: $D_J(A,B) \leq D_J(A,C) + D_J(C,B) $
\end{itemize}

For example, the sets $A = \{ 0, 2, 3, 4, 1 \}$ and $B = \{ 1, 3, 5 \}$ have $D_J(A,B) = \frac{2}{6}$, as their intersection is of cardinality 2 and their union of cardinality 6.
